\index{HolSmtLib|(}
\index{SMT solvers|see {HolSmtLib}}
\index{decision procedures!SMT}

\setcounter{sessioncount}{0}

The purpose of \ml{HolSmtLib} is to provide a platform for
experimenting with combinations of interactive theorem proving and
Satisfiability Modulo Theories~(SMT) solvers.  \ml{HolSmtLib} was
developed as part of a research project on {\it Expressive
  Multi-theory Reasoning for Interactive Verification} (EPSRC grant
EP/F067909/1) from 2008 to~2011.  It is loosely inspired by
\ml{HolSatLib} (Section~\ref{sec:HolSatLib}), and has been described
in parts in the following publications:
\begin{itemize}
\item Tjark Weber: {\it SMT Solvers: New Oracles for the HOL Theorem
  Prover}.  To appear in International Journal on Software Tools for
  Technology Transfer (STTT), 2011.
\item Sascha B{\"o}hme, Tjark Weber: {\it Fast LCF-Style Proof
  Reconstruction for Z3}.  In Matt Kaufmann and Lawrence C.\ Paulson,
  editors, Interactive Theorem Proving, First International
  Conference, ITP 2010, Edinburgh, UK, July 11--14, 2010.
  Proceedings, volume 6172 of Lecture Notes in Computer Science, pages
  179--194.  Springer, 2010.
\end{itemize}
\ml{HolSmtLib} uses external SMT solvers to prove instances of SMT
tautologies, \ie, formulas that are provable using (a combination of)
propositional logic, equality reasoning, linear arithmetic on integers
and reals, and decision procedures for bit vectors and arrays.  The
supported fragment of higher-order logic varies with the SMT solver
used, and is discussed in more detail below.  At least for Yices, it
is a superset of the fragment supported by \ml{bossLib.DECIDE} (and
the performance of \ml{HolSmtLib}, especially on big problems, should
be much better).

\subsection{Interface}

The library currently provides four tactics to invoke different SMT
solvers, namely \ml{CVC\_ORACLE\_TAC}, \ml{YICES\_TAC},
\ml{Z3\_ORACLE\_TAC}, and \ml{Z3\_TAC}.  These tactics are defined in
the \ml{HolSmtLib} structure, which is the library's main entry point.
Given a goal~$(\Gamma, \varphi)$ (where $\Gamma$ is a list of
assumptions, and $\varphi$ is the goal's conclusion), each tactic
returns (i)~an empty list of new goals, and (ii)~a validation function
that returns a theorem~$\Gamma' \vdash \varphi$ (with $\Gamma'
\subseteq \Gamma$). These tactics fail if the SMT solver cannot prove
the goal.\footnote{Internally, the goal's assumptions and the
\emph{negated} conclusion are passed to the SMT solver.  If the SMT
solver determines that these formulas are unsatisfiable, then the
(unnegated) conclusion must be provable from the assumptions.}  In
other words, these tactics solve the goal (or fail).  As with other
tactics, \ml{Tactical.TAC\_PROOF} can be used to derive functions of
type \ml{goal -> thm}.

For each tactic, the \ml{HolSmtLib} structure additionally provides a
corresponding function of type \ml{term -> thm}.  These functions are
called \ml{CVC\_ORACLE\_PROVE}, \ml{YICES\_PROVE},
\ml{Z3\_ORACLE\_PROVE}, and \ml{Z3\_PROVE}, respectively.  Applied to
a formula $\varphi$, they return the theorem $\emptyset \vdash
\varphi$ (or fail).

\paragraph{Oracles vs.\ proof reconstruction}

\ml{CVC\_ORACLE\_TAC}, \ml{YICES\_TAC} and \ml{Z3\_ORACLE\_TAC} use
the SMT solver (cvc5, Yices and Z3, respectively) as an oracle: the
solver's result is trusted.  Bugs in the SMT solver or in
\ml{HolSmtLib} could potentially lead to inconsistent theorems.
Accordingly, the derived theorem is tagged with an oracle tag.

\ml{Z3\_TAC}, on the other hand, performs proof reconstruction.  It
requests a detailed proof from Z3, which is then checked in \HOL{}.
One obtains a proper \HOL{} theorem; no (additional) oracle tags are
introduced. However, Z3's proofs do not always contain enough
information to allow efficient checking in \HOL{}; therefore, proof
reconstruction may be slow or fail. Furthermore, proof reconstruction
only works with Z3~2.19. Newer versions of Z3 are only supported as an
oracle, currently.

\paragraph{Supported subsets of higher-order logic}

\ml{YICES\_TAC} employs a translation into Yices's native input
format.  The interface supports types \holtxt{bool}, \holtxt{num},
\holtxt{int}, \holtxt{real}, \holtxt{->} (\ie, function types),
\holtxt{prod} (\ie, tuples), fixed-width word types, inductive data
types, records, and the following terms: equality, Boolean connectives
(\holtxt{T}, \holtxt{F}, \holtxt{==>}, \holtxt{/\bs}, \holtxt{\bs /},
negation, \holtxt{if-then-else}, \holtxt{bool-case}), quantifiers
(\holtxt{!}, \holtxt{?}), numeric literals, arithmetic operators
(\holtxt{SUC}, \holtxt{+}, \holtxt{-}, \holtxt{*}, \holtxt{/}, unary
minus, \holtxt{DIV}, \holtxt{MOD}, \holtxt{ABS}, \holtxt{MIN},
\holtxt{MAX}), comparison operators (\holtxt{<}, \holtxt{<=},
\holtxt{>}, \holtxt{>=}, both on \holtxt{num}, \holtxt{int}, and
\holtxt{real}), function application, lambda abstraction, tuple
selectors \holtxt{FST} and \holtxt{SND}, and various word operations.

cvc5 and Z3 are integrated via a more restrictive translation into
SMT-LIB~2 format, described below.  Therefore, Yices is typically the
solver of choice at the moment (unless you need proof reconstruction,
which is available for Z3 only).  However, there are a few operations
(\eg, specific word operations) that are supported by the SMT-LIB
format, but not by Yices.  See \ml{selftest.sml} for further details.

Terms of higher-order logic that are not supported by the respective
target solver/\allowbreak translation are typically treated in one of
three ways:
\begin{enumerate}
\item Some unsupported terms are replaced by equivalent suppported
  terms during a pre-processing step.  For instance, all tactics first
  generalize the goal's conclusion by stripping outermost universal
  quantifiers, and attempt to eliminate certain set expressions by
  rewriting them into predicate applications: \eg, \holtxt{y IN \{x |
    P x\}} is replaced by \holtxt{P y}.  The resulting term is
  $\beta$-normalized.  Depending on the target solver, further
  simplifications are performed.
\item Remaining unsupported constants are treated as uninterpreted,
  \ie, replaced by fresh variables.  This should not affect soundness,
  but it may render goals unprovable and lead to spurious
  counterexamples.  To see all fresh variables introduced by the
  translation, you can set \ml{HolSmtLib}'s tracing variable (see
  below) to a sufficiently high value.
\item Various syntactic side conditions are currently not enforced by
  the translation and may result in invalid input to the SMT solver.
  For instance, Yices only allows \emph{linear} arithmetic (\ie,
  multiplication by constants) and word-shifts by numeric literals
  (constants).  If the goal is outside the allowed syntactic fragment,
  the SMT solver will typically fail to decide the problem.
  \ml{HolSmtLib} at present only provides a generic error message in
  this case.  Inspecting the SMT solver's output might provide further
  hints.
\end{enumerate}

\begin{session}
\begin{verbatim}
- load "HolSmtLib"; open HolSmtLib;
(* output omitted *)
> val it = () : unit

- show_tags := true;
> val it = () : unit

- CVC_ORACLE_PROVE ``(a ==> b) /\ (b ==> a) <=> (a <=> b)``;
> val it = [oracles: DISK_THM, HolSmtLib] [axioms: ] []
           |- (a ==> b) /\ (b ==> a) <=> (a <=> b) : thm

- YICES_PROVE ``(a ==> b) /\ (b ==> a) <=> (a <=> b)``;
> val it = [oracles: DISK_THM, HolSmtLib] [axioms: ] []
           |- (a ==> b) /\ (b ==> a) <=> (a <=> b) : thm

- Z3_ORACLE_PROVE ``(a ==> b) /\ (b ==> a) <=> (a <=> b)``;
> val it = [oracles: DISK_THM, HolSmtLib] [axioms: ] []
           |- (a ==> b) /\ (b ==> a) <=> (a <=> b) : thm

- Z3_PROVE ``(a ==> b) /\ (b ==> a) <=> (a <=> b)``;
> val it = [oracles: DISK_THM] [axioms: ] []
           |- (a ==> b) /\ (b ==> a) <=> (a <=> b) : thm
\end{verbatim}
\end{session}

\paragraph{Support for the SMT-LIB 2 file format}

SMT-LIB (see \url{https://smtlib.cs.uiowa.edu/}) is the standard input
format for SMT solvers.  \ml{HolSmtLib} supports (a subset of)
version~2.0 of this format.  A translation of \HOL{} terms into
SMT-LIB~2 format is available in \ml{SmtLib.sml}, and a parser for
SMT-LIB~2 files (translating them into \HOL{} types, terms, and
formulas) can be found in \ml{SmtLib\_Parser.sml}, with auxiliary
functions in \ml{SmtLib\_\{Logics,Theories\}.sml}.

The SMT-LIB~2 translation supports types \holtxt{bool}, \holtxt{int}
and \holtxt{real}, fixed-width word types, and the following terms:
equality, Boolean connectives, quantifiers, numeric literals,
arithmetic operators, comparison operators, function application, and
various word operations.  Notably, the SMT-LIB interface does
\emph{not} support type \holtxt{num}, data types or records, and
higher-order formulas.  See the files mentioned above and the examples
in \ml{selftest.sml} for further details.

\paragraph{Tracing}

Tracing output can be controlled via \ml{Feedback.set\_trace
  "HolSmtLib"}.  See the source code in \ml{Library.sml} for
possible values.

Communication between \HOL{} and external SMT solvers is via temporary
files.  These files are located in the standard temporary directory,
typically {\tt /tmp} on Unix machines.  The actual file names are
generated at run-time, and can be shown by setting the above tracing
variable to a sufficiently high value.

The default behavior of \ml{HolSmtLib} is to delete temporary files
after successful invocation of the SMT solver.  This also can be
changed via the above tracing variable.  If there is an error, files
are retained in any case (but note that the operating system may
delete temporary files automatically, \eg, when \HOL{} exits).

\subsection{Installing SMT solvers}

\ml{HolSmtLib} has been tested with cvc5~1.0.8, Yices~1.0.29, Z3~2.19
and Z3~4.8.17. Later versions may or may not work.  (Yices~2 is not
supported.)  To use \ml{HolSmtLib}, you need to install at least one
of these SMT solvers on your machine.  As mentioned before, Yices
supports a larger fragment of higher-order logic than Z3, but proof
reconstruction has been implemented only for Z3~2.19.

cvc5 is available for various platforms from
\url{https://cvc5.github.io/}. After installation, you must set the
environment variable {\tt \$HOL4\_CVC\_EXECUTABLE} to the pathname of
the cvc5 executable, \eg, {tt /usr/bin/cvc5}, before you invoke \HOL.

Yices is available for various platforms from
\url{https://yices.csl.sri.com/}.  After installation, you must set
the environment variable {\tt \$HOL4\_YICES\_EXECUTABLE} to the
pathname of the Yices executable, \eg, {\tt /bin/yices}, before you
invoke \HOL.

Precompiled binaries of Z3~2.19 for Linux can be downloaded from
\url{https://github.com/Z3Prover/bin/tree/master/legacy}. Note that
Z3~2.19.1 apparently cannot parse real literals (even though Z3~2.19
can), so that version is not recommended since it won't work with
reals. After installation, you must set the environment variable {\tt
\$HOL4\_Z3\_EXECUTABLE} to the pathname of the Z3 executable, \eg,
{\tt /bin/z3}, before you invoke \HOL.

Newer versions of Z3 are available at
\url{https://github.com/Z3Prover/z3}. After installation, you must set
the environment variable {\tt \$HOL4\_Z3\_EXECUTABLE} to the pathname
of the Z3 executable, \eg, {\tt /bin/z3}, before you invoke \HOL.

It should be relatively straightforward to integrate other SMT solvers
that support the SMT-LIB~2 input format as oracles.  However, this
will involve a (typically small) amount of Standard ML programming,
\eg, to interpret the solver's output.  See \ml{CVC.sml} and
\ml{Z3.sml} for some relevant code.

\subsection{Wishlist}

The following features have not been implemented yet.  Please submit
additional feature requests (or code contributions) via
\url{https://github.com/HOL-Theorem-Prover/HOL}.

\paragraph{Counterexamples}

For satisfiable input formulas, SMT solvers typically return a
satisfying assignment.  This assignment could be displayed to the
\HOL{} user as a counterexample.  It could also be turned into a
theorem, similar to the way \ml{HolSatLib} treats satisfying
assignments.

\paragraph{Proof reconstruction}

Proof reconstruction has been implemented only for Z3~2.19. It would
be desirable to implement it for newer versions of Z3, since Z3~2.19
is quite old and is only available as a binary.

Several other SMT solvers can also produce proofs, and it would be
nice to offer \HOL{} users more choice.  However, in the absence of a
standard proof format for SMT solvers, it is perhaps not worth the
implementation effort.

\paragraph{Support for cvc5's and Z3's SMT-LIB extensions}

cvc5 and Z3 support extensions of the SMT-LIB language, \eg, data
types, sets, multisets/bags and strings. \ml{HolSmtLib} does not
utilize these extensions yet when calling these SMT solvers. This
would require the translation for these solvers to be distinct from
the generic SMT-LIB translation.

\paragraph{SMT solvers as a web service}

The need to install an SMT solver locally poses an entry barrier.  It
would be much more convenient to have a web server running one (or
several) SMT solvers, roughly similar to the ``System on TPTP''
interface that G.~Sutcliffe provides for first-order theorem provers
(\url{http://www.cs.miami.edu/~tptp/cgi-bin/SystemOnTPTP}).  For
Isabelle/HOL, such a web service has been installed by S.~B{\"o}hme in
Munich, but unfortunately it is not publicly available.  Perhaps the
SMT-EXEC initiative (\url{http://www.smtexec.org/}) could offer
hardware or implementation support.

\index{HolSmtLib|)}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "description"
%%% End:
